

Loading in packages and making a collected dataframe, where the duplicates within the clusters have been removed
```{r}
library('dplyr')
library('tidyr')
library('tidyverse')
library('ggplot2')
```


```{r}
GeneClustering <- read.csv("New order uTSS mm10 quant FPKM  - 1kb quantile corrected 10c.txt", sep = "\t")
GeneAnnotations <- read.csv("dbtmee_cluster_gene_v2.tsv", sep = "\t")
GeneAnnotations <- as.data.frame(GeneAnnotations)

#removing duplicates from within the clusters
GeneClustering_filtered <- GeneClustering %>% distinct(Group.order, Gene.Name, .keep_all = TRUE)
names(GeneAnnotations)[names(GeneAnnotations) == 'Gene'] <- 'Gene.Name'


head(GeneAnnotations)
head(GeneClustering_filtered)
```


Here there are around 10k genes that are filtered out as there is no annotation for them
```{r}
mergedDF <- merge(GeneClustering_filtered, GeneAnnotations, by = "Gene.Name", all = FALSE)
#mergedDF <- merge(GeneClustering_filtered, GeneAnnotations, by = "Gene.Name", all = TRUE)
```

Making a new dataframe only keeping the columns needed and then renaming them to fit the data
(classification, cluster, group1, group2, group3, etc.)
```{r}
 df <- mergedDF %>% select("Cluster.y", "Order.based.on.stages", "Gene.Name", "X6.p7.Az.250.16_S14_R1_001", "X4.p10B.AZ.250.14_S10_R1_001", "X1.p12.AZ.200_S1_L001_R1_001_pooled", "X3.NSN.AZ.200_S3_L001_R1_001_pooled", "X4.SN.AZ.200_S4_L001_R1_001_pooled", "X5.MII.AZ.300_S5_L001_R1_001_pooled", "X8.8c.AZ.460_S8_L001_R1_001", "X11.mor.AZ460_S11_L001_R1_001", "X13.bla.AZ460_S13_L001_R1_001")

df <- setNames(df, c("Gene_type","Cluster","Gene_name", "P7", "P10", "p12", "NSN", "SN", "MII", "8cell", "Morula", "Blastocyst"))

#names(mergedDF)
```

The log2fold calculation

Getting the clusters and total counts for the genetypes
```{r}
allClusters <- 1:(nrow(table(df$Cluster)))
totalcounts1 <- as.data.frame(table(df$Gene_type))
totalcounts <- as.numeric(totalcounts1$Freq)
allInsertions <- as.character(totalcounts1$Var1)
Genetypes <- allInsertions
#allInsertions <- Genetypes

df1 <- df %>%  mutate(value_col = 1) %>% pivot_wider(names_from = Gene_type, values_from = value_col, values_fill = 0)
```


```{r}
#initiating matrix
globaldf <- data.frame(x=allInsertions, counts=totalcounts)
finalmatrix <- matrix(0, nrow = length(allInsertions), ncol = length(allClusters))
rownames(finalmatrix) <- allInsertions
colnames(finalmatrix) <- allClusters
```


```{r}
#creating nested loop as to make a table with cluster/logfold for chosen strain
for (GENEx in Genetypes) {
  for (clusterx in allClusters) {
    cluster1 <- df1 %>% 
      select('Cluster', Genetypes) %>% filter(Cluster == clusterx )
    
    counts <- sapply(allInsertions, function(col) sum(cluster1[[col]], na.rm = FALSE))
    localdf <- data.frame(x=allInsertions, counts=counts)
    localdf <- localdf %>% mutate(across(where(is.numeric), ~ .x + 1)) #adding 1 to every value to avoid 0
    globaldf$freq <- globaldf$counts/sum(globaldf$counts)
    totallocalcounts <- sum(localdf$counts)
    localdf$expectedcounts <- totallocalcounts*globaldf$freq
    localdf$logfoldchange <- log2(localdf$counts/localdf$expectedcounts)
    
    finalmatrix[GENEx, as.character(clusterx)] <- localdf$logfoldchange[which(localdf$x == GENEx)]
    
  }
}

finaldf <- as.data.frame(as.table(finalmatrix))

# Rename the columns for clarity
colnames(finaldf) <- c("Genex", "Clusterx", "Log2foldchange")

```




The statistical analysis 
Warnings are silenced, as there will be warnings where it finds a 0.
```{r, warning=FALSE}
table <- df1


#defining the chi square test function
perform_chi_square_test <- function(row) {
  matrix_data <- matrix(c(row$INSin, row$INSout, row$nonINSin, row$nonINSout), nrow = 2, byrow = TRUE)
  test_result <- chisq.test(matrix_data, correct = FALSE)
  return(test_result)
}

finalmatrix <- matrix(NA, nrow = length(allClusters), ncol = length(Genetypes))
colnames(finalmatrix) <- Genetypes
rownames(finalmatrix) <- allClusters


for (INSx in Genetypes) {
  list1 <- c()
  list2 <- c()
  list3 <- c()
  list4 <- c()
  for (clusterx in allClusters) {
    
    INSin <- table %>% filter(.data[[INSx]] == 0) %>% filter(Cluster == clusterx) 
    INSout <- table %>% filter(.data[[INSx]] == 1) %>% filter(Cluster == clusterx) 
    nonINSin <- table %>% filter(.data[[INSx]] == 0) %>% filter(Cluster != clusterx) 
    nonINSout <- table %>% filter(.data[[INSx]] == 1) %>% filter(Cluster != clusterx)
    list1<- append(list1, nrow(INSin))
    list2<- append(list2, nrow(INSout))
    list3<- append(list3, nrow(nonINSin))
    list4<- append(list4, nrow(nonINSout))
    
    df <- data.frame(list2, list4, list1, list3)
    names(df) <- c('INSin','INSout','nonINSin','nonINSout')
    df2 <- tibble(df)
    
    chi_square_results <- df2 %>% 
      rowwise() %>% 
      mutate(
        test_result = list(perform_chi_square_test(cur_data()))
      ) %>% 
      ungroup() %>% 
      mutate(
        p_value = map_dbl(test_result, ~ .x$p.value)
      ) %>% 
      select(-test_result)
    
    pval <- chi_square_results$p_value
    p_adjusted <- p.adjust(pval, method = 'BH')
    chi_square_results$p_adjusted <- p_adjusted
    chi_square_results$x <- as.numeric(row.names(chi_square_results))
    
    finalmatrix[clusterx, as.character(INSx)] <- chi_square_results$p_value[which(chi_square_results$x == clusterx)]
    
  }
}

# matrix with adjusted p-values:
p_vector <- as.vector(finalmatrix)

# apply the p-value adjustment (Benjamini-Hochberg)
p_adjusted_vector <- p.adjust(p_vector, method = "BH")

# reshaping the adjusted p-values vector back into the original matrix dimensions
adjusted_p_matrix <- matrix(p_adjusted_vector, nrow = nrow(finalmatrix), ncol = ncol(finalmatrix))
colnames(adjusted_p_matrix) <- allInsertions
rownames(adjusted_p_matrix) <- allClusters

finaldf2 <- as.data.frame(as.table(adjusted_p_matrix))

# Rename the columns for clarity
colnames(finaldf2) <- c("Clusterx2", "Genex", "pvalue")

```


The order of the gene types for plotting
```{r}
#We want it to be this order
Genetypes <- c("Maternal RNA",  "Maternal to ZGA", "Minor ZGA", "1-Cell Transient", "Major ZGA", "2-Cell Transient", "Maternal to MGA", "Minor ZGA to MGA", "MGA", "4-Cell Transient")
```


Merging data and creating plots
```{r}
finaldf2$ID <- paste(finaldf2$Clusterx, finaldf2$Genex)
finaldf$ID <- paste(finaldf$Clusterx, finaldf$Genex)

df_merge <- merge(finaldf, finaldf2,by="ID") 

colnames(df_merge)[colnames(df_merge) == 'Adjusted P-value'] <- 'pvalue'

# Modify the significance column to numeric values
df_merge$logsignificance <- -log10(df_merge$pvalue)
df_merge$Clusterx = as.numeric(df_merge$Clusterx)
```


Modifying the maximum logfold and -log10Padjusted values to fit the plot
```{r}
df_merge$Log2foldchange <- with(df_merge, ifelse(Log2foldchange > 1, 1, ifelse(Log2foldchange < -1, -1, Log2foldchange)))
df_merge$logsignificance <- with(df_merge, ifelse(logsignificance > 20, 20, logsignificance))

# defining the significance from the -log10 p value calculations
oursignificance <- -log10(0.05)
```


```{r}
# Plot using the modified significance column
a <- ggplot(df_merge, aes(y = Clusterx, x = factor(Genex.x, levels = Genetypes), size = logsignificance)) +
  geom_point(aes(color = Log2foldchange)) +
  theme_minimal() +
  labs(title = paste("Differntial enrichment analysis for clustering based on TSSs"),
       x = "Gene type",
       y = "Cluster") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#  scale_size_continuous(range = c(1, 10)) +             # Adjust the range to control the size of the points for only one plot
  scale_y_reverse(breaks=1:20) +
  scale_size(range = c(1, 15), limits = c(oursignificance,20)) +  # this used to compare between plots. to make them comparable, range decides size and limits decides top value
  scale_color_gradient2(low = "#77008f", mid = "white", high = "#018d1f", midpoint = 0,limits = c(-1.01, 1.01)) #Limit sets the top and bottom, but values that falls outside this category will be grey


a



```



