

Loading in packages and making a collected dataframe, where the duplicates within the clusters have been removed
```{r}
library('dplyr')
library('tidyr')
library('tidyverse')
library('ggplot2')
library("readxl")
```


```{r}
df_excel <- read_excel("Peaks no TSS or CpG repeatmasker.xlsx")
```


```{r}
df <- df_excel

# Removing the additional added numbers
for ( col in 1:ncol(df)){
    colnames(df)[col] <-  sub("\\...*", "", colnames(df)[col])
}

df <- df[1:(nrow(df)-1),] # removing sum column

names(df)[names(df) == 'Order based on stages'] <- 'Cluster'

types_class <- colnames(df[12:19])

types_family <- colnames(df[21:39]) # change the names

types_name <- colnames(df[41:130]) # change the names
```

Making a new dataframe only keeping the columns needed and then renaming them to fit the data
(classification, cluster, group1, group2, group3, etc.)
```{r}
df_class <- df %>% select("Cluster", types_class)

#df_class <- setNames(df_class, c("Cluster", "SINE", "LTR", "Simple_repeat", "LINE", "Low_complexity", "DNA", "Other", "Satellite"))

df_family <- df %>% select("Cluster", types_family)

df_name <- df %>% select("Cluster", types_name)

```


Initiating the process with the data and parameters:

chosen classification to make the type distinction
```{r}
df <- df_class
types <- types_class
```

The log2fold calculation

Getting the clusters and total counts for the types
```{r}
allClusters <- 1:(nrow(table(df$Cluster)))

totalcounts <- sapply(types, function(col) sum(df[[col]], na.rm = FALSE))
sum(totalcounts)
```

Removing zeros/(maybe not nescessary but implement the rows with no type)
```{r}
df_no_cluster <- df %>% select(types)
zero_rows <- df_no_cluster[rowSums(df_no_cluster == 0) == ncol(df_no_cluster), ]

```


```{r}
totalcounts <- sapply(types, function(col) sum(df[[col]], na.rm = FALSE))
hej <- as.data.frame(totalcounts)


```


```{r}
#initiating matrix
globaldf <- data.frame(x=types, counts=totalcounts)
finalmatrix <- matrix(0, nrow = length(types), ncol = length(allClusters))
rownames(finalmatrix) <- types
colnames(finalmatrix) <- allClusters
```


```{r}
#creating nested loop as to make a table with cluster/logfold for chosen strain
for (GENEx in types) {
  for (clusterx in allClusters) {
    cluster1 <- df %>% 
      select('Cluster', types) %>% filter(Cluster == clusterx )
    
    counts <- sapply(types, function(col) sum(cluster1[[col]], na.rm = FALSE))
    localdf <- data.frame(x=types, counts=counts)
    localdf <- localdf %>% mutate(across(where(is.numeric), ~ .x + 1)) #adding 1 to every value to avoid 0
    globaldf$freq <- globaldf$counts/sum(globaldf$counts)
    totallocalcounts <- sum(localdf$counts)
    localdf$expectedcounts <- totallocalcounts*globaldf$freq
    localdf$logfoldchange <- log2(localdf$counts/localdf$expectedcounts)
    
    finalmatrix[GENEx, as.character(clusterx)] <- localdf$logfoldchange[which(localdf$x == GENEx)]
    
  }
}

finaldf <- as.data.frame(as.table(finalmatrix))

# Rename the columns for clarity
colnames(finaldf) <- c("Genex", "Clusterx", "Log2foldchange")

```




The statistical analysis 
Warnings are silenced, as there will be warnings where it finds a 0.
```{r, warning=FALSE}
table <- df


#defining the chi square test function
perform_chi_square_test <- function(row) {
  matrix_data <- matrix(c(row$INSin, row$INSout, row$nonINSin, row$nonINSout), nrow = 2, byrow = TRUE)
  test_result <- chisq.test(matrix_data, correct = FALSE)
  return(test_result)
}

finalmatrix <- matrix(NA, nrow = length(allClusters), ncol = length(types))
colnames(finalmatrix) <- types
rownames(finalmatrix) <- allClusters


for (INSx in types) {
  list1 <- c()
  list2 <- c()
  list3 <- c()
  list4 <- c()
  for (clusterx in allClusters) {
    
    INSin <- table %>% filter(.data[[INSx]] == 0) %>% filter(Cluster == clusterx) 
    INSout <- table %>% filter(.data[[INSx]] == 1) %>% filter(Cluster == clusterx) 
    nonINSin <- table %>% filter(.data[[INSx]] == 0) %>% filter(Cluster != clusterx) 
    nonINSout <- table %>% filter(.data[[INSx]] == 1) %>% filter(Cluster != clusterx)
    list1<- append(list1, nrow(INSin))
    list2<- append(list2, nrow(INSout))
    list3<- append(list3, nrow(nonINSin))
    list4<- append(list4, nrow(nonINSout))
    
    df <- data.frame(list2, list4, list1, list3)
    names(df) <- c('INSin','INSout','nonINSin','nonINSout')
    df2 <- tibble(df)
    
    chi_square_results <- df2 %>% 
      rowwise() %>% 
      mutate(
        test_result = list(perform_chi_square_test(cur_data()))
      ) %>% 
      ungroup() %>% 
      mutate(
        p_value = map_dbl(test_result, ~ .x$p.value)
      ) %>% 
      select(-test_result)
    
    pval <- chi_square_results$p_value
    p_adjusted <- p.adjust(pval, method = 'BH')
    chi_square_results$p_adjusted <- p_adjusted
    chi_square_results$x <- as.numeric(row.names(chi_square_results))
    
    finalmatrix[clusterx, as.character(INSx)] <- chi_square_results$p_value[which(chi_square_results$x == clusterx)]
    
  }
}

# matrix with adjusted p-values:
p_vector <- as.vector(finalmatrix)

# apply the p-value adjustment (Benjamini-Hochberg)
p_adjusted_vector <- p.adjust(p_vector, method = "BH")

# reshaping the adjusted p-values vector back into the original matrix dimensions
adjusted_p_matrix <- matrix(p_adjusted_vector, nrow = nrow(finalmatrix), ncol = ncol(finalmatrix))
colnames(adjusted_p_matrix) <- types
rownames(adjusted_p_matrix) <- allClusters

finaldf2 <- as.data.frame(as.table(adjusted_p_matrix))

# Rename the columns for clarity
colnames(finaldf2) <- c("Clusterx2", "Genex", "pvalue")

```





Merging data and creating plots
```{r}
finaldf2$ID <- paste(finaldf2$Clusterx, finaldf2$Genex)
finaldf$ID <- paste(finaldf$Clusterx, finaldf$Genex)

df_merge <- merge(finaldf, finaldf2, by="ID") 

colnames(df_merge)[colnames(df_merge) == 'Adjusted P-value'] <- 'pvalue'

# Modify the significance column to numeric values
df_merge$logsignificance <- -log10(df_merge$pvalue)
df_merge$Clusterx = as.numeric(df_merge$Clusterx)

# Modify the numbers for plotting
df_merge$Log2foldchange <- with(df_merge, ifelse(df_merge$Log2foldchange > 1, 1, ifelse(df_merge$Log2foldchange < -1, -1, df_merge$Log2foldchange)))

df_merge$logsignificance <- with(df_merge, ifelse(logsignificance > 20, 20, logsignificance))
```


```{r}
# Plot using the modified significance column
a <- ggplot(df_merge, aes(y = Clusterx, x = factor(Genex.x, levels = types), size = logsignificance)) +
  geom_point(aes(color = Log2foldchange)) +
  theme_minimal() +
  labs(title = paste("clusters and expression of genes"),
       x = "Gene type",
       y = "Cluster") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#  scale_size_continuous(range = c(1, 10)) +             # Adjust the range to control the size of the points for only one plot
  scale_y_reverse(breaks=1:20) +
  scale_size(range = c(1, 15), limits = c(0.1,20)) +  # this used to compare between plots. to make them comparable, range decides size and limits decides top value
  scale_color_gradient2(low = "#77008f", mid = "white", high = "#018d1f", midpoint = 0,limits = c(-1, 1)) #Limit sets the top and bottom, but values that falls outside this category will be grey


a



```

Here sorting the Transposable elements on name level by class(only do this if visualizing on name level):

```{r}
df_types <- read_excel("repeat type frequencies.xlsx")

# Getting the class for each 

df_classification <- df_excel[,8:10]
df_classification <- setNames(df_classification, c("Name", "Class", "Family"))
df_diff_names <- df_classification %>%  filter(Name %in% types_name) %>%  distinct()
#df_diff_names <- df_diff_names[order(df_diff_names$Class, decreasing = FALSE),]
df_diff_names <- arrange(df_diff_names, Class, Family)

# names ordered by class
names_ordered <- df_diff_names$Name


a <- ggplot(df_merge, aes(y = Clusterx, x = factor(Genex.x, levels = names_ordered), size = logsignificance)) +
  geom_point(aes(color = Log2foldchange)) +
  theme_minimal() +
  labs(title = paste("clusters and expression of genes"),
       x = "Gene type",
       y = "Cluster") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  #  scale_size_continuous(range = c(1, 10)) +             # Adjust the range to control the size of the points for only one plot
  scale_y_reverse(breaks=1:20) +
  scale_size(range = c(1, 15), limits = c(0.1,20)) +  # this used to compare between plots. to make them comparable, range decides size and limits decides top value
  scale_color_gradient2(low = "#77008f", mid = "white", high = "#018d1f", midpoint = 0,limits = c(-1, 1)) #Limit sets the top and bottom, but values that falls outside this category will be grey


a

```


```{r}
install.packages('writexl')

library(writexl)

write_xlsx(df_diff_names, "TE_Classicifations.xlsx")
```



















